---
title: "Supplement: Symmetric Case"
output: html_document
date: "2023-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, echo=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
```


We test whether we can use the independence of ranks of the same column element in two different rows.

First, we set up the parameters: number of runs, the size of the matrix, and whether the main diagonal is zero.

```{r parameters}
msize<-10
n<-10000
set.seed(42)
```


# Symmetric square matrix

We make a symmetric random matrix n times and then we save the ranks of 3 row in 1 column and in second column for each permutation.

```{r experiment_sym}
zerodiag<-FALSE
two_columns<-data.table('u31'=rep(-1,n),'u32'=rep(-1,n))
for (i in 1:n) {
  matr<-matrix(runif(msize**2),nrow=msize)
  matr<-matr+t(matr)
  if(zerodiag) {diag(matr)<-0}
  two_columns[i,1]<-msize-rank(matr[,1])[3]
  two_columns[i,2]<-msize-rank(matr[,2])[3]
}

```

The Kendall correlation does not detect any correlation between the vectors.

```{r correlation_sym}
cor.test(two_columns[[1]],two_columns[[2]],method="kendall")
```


Let's plot the results.

```{r plot_sym}
two_columns<-two_columns %>% count (u31,u32,name="counter")
scale<-median(two_columns$counter)/4
dots <- ggplot(two_columns,aes(x=u31,y=u32)) + 
  geom_point(size=two_columns$counter/scale,alpha=0.5,col="blue") + 
  theme_classic()
plot(dots)
```

Let's test the uniformity.
```{r uniformity_sym}
chisq.test(two_columns$counter)
```


# Symmetric square matrix, zero diagonal

```{r experiment_sym_zero_diag}
zerodiag<-TRUE
two_columns<-data.table('u31'=rep(-1,n),'u32'=rep(-1,n))
for (i in 1:n) {
  matr<-matrix(runif(msize**2),nrow=msize)
  matr<-matr+t(matr)
  if(zerodiag) {diag(matr)<-0}
  two_columns[i,1]<-msize-rank(matr[,1])[3]
  two_columns[i,2]<-msize-rank(matr[,2])[3]
}
```


```{r correlation_sym_zero_diag}
cor.test(two_columns[[1]],two_columns[[2]],method="kendall")
```


```{r plot_sym_zero_diag}
two_columns<-two_columns %>% count (u31,u32,name="counter")
scale<-median(two_columns$counter)/4
dots <- ggplot(two_columns,aes(x=u31,y=u32)) + 
  geom_point(size=two_columns$counter/scale,alpha=0.5,col="blue") + 
  theme_classic()
plot(dots)
```

```{r uniformity_sym_zero_diag}
chisq.test(two_columns$counter)
```

# Asymmetric square matrix

Let's run the procedure for a random asymmetric square matrix.

```{r experiment_asym}
zerodiag<-FALSE
two_columns<-data.table('u31'=rep(-1,n),'u32'=rep(-1,n))
for (i in 1:n) {
  matr<-matrix(runif(msize**2),nrow=msize)
  #matr<-matr+t(matr) 
  #symmetriser, commented
  if(zerodiag) {diag(matr)<-0}
  two_columns[i,1]<-msize-rank(matr[,1])[3]
  two_columns[i,2]<-msize-rank(matr[,2])[3]
}
```

The Kendall correlation does not detect any correlation between the vectors.

```{r correlation_asym}
cor.test(two_columns[[1]],two_columns[[2]],method="kendall")
```


Let's plot the results.

```{r plot_asym}
two_columns<-two_columns %>% count (u31,u32,name="counter")
scale<-median(two_columns$counter)/4
dots <- ggplot(two_columns,aes(x=u31,y=u32)) + 
  geom_point(size=two_columns$counter/scale,alpha=0.5,col="blue") + 
  theme_classic()
plot(dots)
```


Let's test the uniformity.
```{r uniformity_asym}
chisq.test(two_columns$counter)
```



# Asymmetric square matrix, zero diagonal

Let's run the procedure for a random asymmetric square matrix.

```{r experiment_asym_zero_diag}
zerodiag<-TRUE
two_columns<-data.table('u31'=rep(-1,n),'u32'=rep(-1,n))
for (i in 1:n) {
  matr<-matrix(runif(msize**2),nrow=msize)
  #matr<-matr+t(matr) 
  #symmetriser, commented
  if(zerodiag) {diag(matr)<-0}
  two_columns[i,1]<-msize-rank(matr[,1])[3]
  two_columns[i,2]<-msize-rank(matr[,2])[3]
}
```

The Kendall correlation does not detect any correlation between the vectors.

```{r correlation_asym_zero_diag}
cor.test(two_columns[[1]],two_columns[[2]],method="kendall")
```


Let's plot the results.

```{r plot_asym_zero_diag}
two_columns<-two_columns %>% count (u31,u32,name="counter")
scale<-median(two_columns$counter)/4
dots <- ggplot(two_columns,aes(x=u31,y=u32)) + 
  geom_point(size=two_columns$counter/scale,alpha=0.5,col="blue") + 
  theme_classic()
plot(dots)
```


Let's test the uniformity.
```{r uniformity_asym_zero_diag}
chisq.test(two_columns$counter)

```



# Nonsquare matrix

For control, let's run the procedure for a nonsquare matrix.

```{r experiment_nonsq}

two_columns<-data.table('u31'=rep(-1,n),'u32'=rep(-1,n))
nsize=2*msize
for (i in 1:n) {
  matr<-matrix(runif(msize*nsize),nrow=msize)
  if(zerodiag) {diag(matr)<-0}
  two_columns[i,1]<-msize-rank(matr[,1])[3]
  two_columns[i,2]<-msize-rank(matr[,2])[3]
}
```

The Kendall correlation does not detect any correlation between the vectors.

```{r correlation_nonsq}
cor.test(two_columns[[1]],two_columns[[2]],method="kendall")
```


Let's plot the results.

```{r plot_asym_nonsq}
two_columns<-two_columns %>% count (u31,u32,name="counter")
scale<-median(two_columns$counter)/4
dots <- ggplot(two_columns,aes(x=u31,y=u32)) + 
  geom_point(size=two_columns$counter/scale,alpha=0.5,col="blue") + 
  theme_classic()
plot(dots)
```


Let's test the uniformity.
```{r uniformity_nonsq}
chisq.test(two_columns$counter)
```

